Dependently Typed Languages in Statix

### 2022-10-26 paper-evcs/paper.tex
Process feedback on paper

Dependently typed languages are different from other languages, because they allow types to be parameterized by values. This allows more rigorous reasoning over types and the values that are inhabited by a type. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires evaluation of the terms they are parameterized by.

This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language. We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix. The goal is not to show that Statix can implement it, but that implementing it is easier in Statix than in a general-purpose programming language.

We will first show the base language and explain the way that Statix was used to implement this language. Next, we will explore several features and see how well they can be expressed in Statix.

The implementation of the language is available on GitHub. The \verb|final-simple| branch contains the implementation that is relevant for this paper. \\
\url{https://github.com/JonathanBrouwer/master-thesis/}

@@

\subsection{The language}

@@

\subsection{Scope Graphs}

To type-check the base language, we need to store information about the names we have encountered. There are two different situations, names that we encounter that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}

Both of these need to be stored in the scope graph. The scope graph only has a single type of edge, called \verb|P| (parent) edges. It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.

@@

\newpage
\subsection{Beta Reductions}

A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare. Beta reduction is done using a Krivine abstract machine\cite{krivine}.

@@

		\inferrule{ 
			sGetName(s, n) = NSubst(se, e) 
			\\ se \vdash e, t \bhr se', e' 
		}{ s \vdash Var(n), t \bhr se', e' }

@@

\newpage
\subsection{Type-checking the Calculus of Constructions}

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type-checking rules for the language. We use some syntax in order to make the rules a bit more compact. 
\begin{enumerate}
	\item $s \vdash e : t$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $s1 \vdash e1 \underset{\beta}{=} s2 \vdash e2$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $s1 \vdash e1 \bhr s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
	\item $s1 \vdash e1 \underset{\beta}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

\newpage
\section{Extensions}
This section will discuss some further ideas that can be explored to build upon what has already been shown. These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.

\paragraph*{Avoiding variable capture}
One problem with the current implementation is that it does not avoid variable capture. Variable capture happens when a term becomes bound because of a wrong substitution. For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type. \T : T. T| is \verb|T : Type -> T : T -> T|. This is not the correct type, and even worse, the correct type is not possible to express without renaming! The problem is that there are multiple names, and there is no way to distinguish between them. This can be solved by using scopes to distinguish names.

\paragraph*{Term Inference}
In dependently typed language, we can infer the value of types. Ideally, we would like to infer the most general type possible. However, this kind of analysis is not possible in Statix because you cannot reason over whether meta-variables are instantiated or not. We implemented an approximation to inference that can infer most types.

\paragraph*{Inductive Data Types}
Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types. Both of these are also fully possible to express in Statix.

@@

\bibliography{oasics-v2021-sample-article}

>>>>

Dependently typed languages are different from other languages, because they allow types to be parameterized by values. This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires evaluation of the terms they are parameterized by.

This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language. We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix. The goal is not to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.

In this paper we show that Statix is already usable as a tool for implementing dependent types. There were some challenges in implementing the dependent type system, which this paper also discusses. Finally, we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.

The implementation of the language is available on GitHub.\\
\url{https://github.com/JonathanBrouwer/master-thesis/}

@@

In this section, we will describe how to implement a dependently typed language in Statix. In section \ref{sec:coc-syntax} we will describe the syntax of the language, then in section \ref{sec:coc-scopes} we will describe how scope graphs are used to type check the language. Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type-check the language.

\subsection{The language}
\label{sec:coc-syntax}

@@

An example program is the following, which defines a polymorphic identity function and applies it to a function:

\begin{lstlisting}
let f = \T: Type. \x: T. x;
f (_: Type -> Type) (\x: Type. x)
\end{lstlisting}
The AST of this program would be:
\begin{lstlisting}
Let(
  "f", 
  FnConstruct("T", Type(), FnConstruct("x", Var("T"), Var("x"))),
  FnDestruct(
	FnDestruct(Var("f"), FnConstruct("_", Type(), Type())),
	FnConstruct("x", Type(), Var("x"))
  )
)
\end{lstlisting}

\subsection{Scope Graphs}
\label{sec:coc-scopes}

To type-check the base language, we need to store information about the names that are in scope at each point in the program. There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}

In Statix, all this information can be stored in a \verb|scope graph|\cite{scope_graphs}, which is a feature of Statix. We only use a single type of edge, called \verb|P| (parent) edges. It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.

@@

\subsection{Beta Reductions}
\label{sec:coc-dynsyms}

A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare.

We implement beta reduction using a Krivine abstract machine\cite{krivine}. This turned out to be the more natural way of expressing this over substitution-based evaluation relation. We originally tried to implement a substitution-based evaluation relation, which works fine for the base language. However, it runs into trouble when implementing inductive data types, more information about this will be in the full master thesis. Furthermore, abstract machines are usually more efficient than substitution-based approaches.

Note that the evaluator uses names rather than De Bruijn indices. In conventional dependently typed languages, evaluation is often done using De Bruijn indices. However, since the aim of this paper is to use the features of Statix, and scope graphs work based on names, we chose to use names.

@@

		\inferrule{ 
			sGetName(s, n) = NSubst(se, e) 
			\\ se \vdash e, t \bhr se' \vdash  e' 
		}{ s \vdash Var(n), t \bhr se' \vdash e' }

@@

\subsection{Type-checking the Calculus of Constructions}
\label{sec:coc-typecheck}

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type-checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
\begin{enumerate}
	\item $s \vdash e : t$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $s1 \vdash e1 \underset{\beta}{=} s2 \vdash e2$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $s1 \vdash e1 \bhr s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
	\item $s1 \vdash e1 \underset{\beta}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

\subsection{Avoiding variable capture}
One problem with the current implementation is that it does not avoid variable capture. Variable capture happens when a term becomes bound because of a wrong substitution. For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type. \T : T. T| is \verb|T : Type -> T : T -> T|. This is not the correct type, and even worse, the correct type is not possible to express without renaming! The problem is that there are multiple names, and there is no way to distinguish between them. This can be solved by using scopes to distinguish names.

\section{Extensions}
This section will discuss some further ideas that can be explored to build upon what has already been shown. These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.

\paragraph*{Term Inference}
In dependently typed language, we can infer the value of types. Ideally, we would like to infer the most general type possible. However, this kind of analysis is not possible in Statix because you cannot reason over whether meta-variables are instantiated or not. We implemented an approximation to inference that can infer most types.
\begin{lstlisting}
(\x: _. x) true
\end{lstlisting}

\paragraph*{Inductive Data Types}
Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types. We can also generate eliminators for these data types to use their values. All of this has been implemented in Statix. 
\begin{lstlisting}
data Maybe (T : Type) = 
    None : Maybe T,
    Some : _: T -> Maybe T;
\end{lstlisting}

@@

\bibliography{paper}
